---
title: "Assignment 3 - Aliens"
author: "Study Group 6"
date: "2026-04-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup
```{r}
# Loading required packages
pacman::p_load(
  tidyverse,    # Data manipulation and plotting
  here,         # Robust file paths
  cmdstanr,     # Stan interface
  posterior,    # Posterior analysis
  bayesplot,    # Plotting MCMC
  furrr,        # Parallel processing
  future,       # Future backend
  here,
  loo
)
```

## Load in empirical data
```{r}
df <- read.csv(here("advanced_cognitive_modeling/Assignment_4/AlienData.txt"))
```

## Preprocessing

```{r}
# remove pt trials
df <- df %>%
  filter(test != 1 & session == 1)

# extract feature names (first session: danger is feature 1+3=1; nutrition is feature 4=1)
df$feature1 <- as.numeric(substr(df$stimulus, 1, nchar(df$stimulus) - 8))
df$feature2 <- as.numeric(substr(df$stimulus, 2, nchar(df$stimulus) - 7))
df$feature3 <- as.numeric(substr(df$stimulus, 3, nchar(df$stimulus) - 6))
df$feature4 <- as.numeric(substr(df$stimulus, 4, nchar(df$stimulus) - 5))
df$feature5 <- as.numeric(substr(df$stimulus, 5, nchar(df$stimulus) - 4))

# make response_dangerous feature
df$response_danger <- ifelse(df$response == 3 | df$response == 4, 1, 0)
```


```{r}
head(df)
names(df)
```

## Functions

```{r}
# Distance calculation: City-block distance (r = 1), attention-weighted
distance <- function(vect1, vect2, w) {
  sum(w * abs(vect1 - vect2))
}

# Similarity computation: Exponential similarity kernel
similarity <- function(dist_val, c) {
  exp(-c * dist_val)
}
```


## GCM model
```{r}
# Generative GCM model function.
gcm_simulate <- function(w,            # attention weights (sums to 1)
                         c,            # sensitivity (natural scale, strictly positive)
                         bias,         # response bias towards category 1
                         obs,          # matrix of stimulus features (trials x features)
                         cat_feedback, # true category labels (0 or 1)
                         quiet = TRUE) {
  ntrials   <- nrow(obs)
  nfeatures <- ncol(obs)

  prob_cat1    <- numeric(ntrials)
  sim_response <- numeric(ntrials)

  memory_obs <- matrix(NA_real_, nrow = 0, ncol = nfeatures)
  memory_cat <- numeric(0)

  for (i in seq_len(ntrials)) {
    if (!quiet && i %% 20 == 0) cat("Simulating trial:", i, "\n")

    current_stim <- as.numeric(obs[i, ])
    n_mem        <- nrow(memory_obs)
    has_cat0     <- any(memory_cat == 0)
    has_cat1     <- any(memory_cat == 1)

    if (n_mem == 0 || !has_cat0 || !has_cat1) {
      prob_cat1[i] <- bias
    } else {
      sims <- vapply(seq_len(n_mem), function(e) {
        similarity(distance(current_stim, memory_obs[e, ], w), c)
      }, numeric(1))

      # Mean similarity per category — removes category-frequency bias.
      m1  <- mean(sims[memory_cat == 1])
      m0  <- mean(sims[memory_cat == 0])
      num <- bias * m1
      den <- num + (1 - bias) * m0

      prob_cat1[i] <- if (den > 1e-9) num / den else bias
      prob_cat1[i] <- pmax(1e-9, pmin(1 - 1e-9, prob_cat1[i]))
    }

    sim_response[i] <- rbinom(1, 1, prob_cat1[i])

    memory_obs <- rbind(memory_obs, current_stim)
    memory_cat <- c(memory_cat, cat_feedback[i])
  }

  tibble(prob_cat1 = prob_cat1, sim_response = sim_response)
}
```


```{r}
stimulus_info <- df %>%
  select(stimulus, feature1, feature2, feature3, feature4, feature5, dangerous) %>%
  unique() %>%
  rename(category_true = dangerous)
```

```{r}
n_blocks         <- 3
n_stim_per_block <- nrow(stimulus_info)
total_trials     <- n_blocks * n_stim_per_block
```


```{r}
make_subject_schedule <- function(stimulus_info, n_blocks, seed) {
  set.seed(seed)
  n_stim <- nrow(stimulus_info)

  sequence <- unlist(lapply(seq_len(n_blocks), function(b) {
    sample(stimulus_info$stimulus, n_stim, replace = FALSE)
  }))

  tibble(
    trial_within_subject = seq_along(sequence),
    block                = rep(seq_len(n_blocks), each = n_stim),
    stimulus_id          = sequence
  ) |>
    left_join(stimulus_info, by = c("stimulus_id" = "stimulus")) |>
    rename(category_feedback = category_true) |>
    dplyr::select(trial_within_subject, block, stimulus_id,
                  feature1, feature2, feature3, feature4, feature5, category_feedback)
}
```


```{r}
simulate_gcm_agent <- function(agent_id, w_true, c_true, bias_true,
                               stimulus_info, n_blocks, subject_seed) {
  schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = subject_seed)
  obs      <- schedule |> dplyr::select(feature1, feature2, feature3, feature4, feature5)
  feedback <- schedule$category_feedback

  sim_results <- gcm_simulate(
    w = w_true, c = c_true, bias = bias_true,
    obs = obs, cat_feedback = feedback, quiet = TRUE
  )

  schedule |>
    mutate(
      agent_id     = agent_id,
      w1_true      = w_true[1],
      w2_true      = w_true[2],
      w3_true      = w_true[3],
      w4_true      = w_true[4],
      w5_true      = w_true[5],
      c_true       = c_true,
      log_c_true   = log(c_true),
      bias_true    = bias_true,
      prob_cat1    = sim_results$prob_cat1,
      sim_response = sim_results$sim_response,
      correct      = as.integer(category_feedback == sim_response)
    ) |>
    group_by(agent_id) |>
    mutate(performance = cumsum(correct) / row_number()) |>
    ungroup()
}
```

```{r}
regenerate_simulations = TRUE
```


```{r}
param_grid <- expand_grid(
  w_setting    = list(c(0.2, 0.2, 0.2, 0.2, 0.2), c(0.1, 0.1, 0.1, 0.6, 0.1), c(0.25, 0.25, 0.25, 0.05, 0.2)), # 3 scenarios
  c_setting    = c(0.5, 1.0, 2.0, 4.0, 5.0, 15.0), # 6 scenarios
  bias_setting = 0.5
) |>
  mutate(
    setting_id = row_number(),
    w_label    = map_chr(w_setting, ~paste(round(.x, 1), collapse = ",")),
    c_label    = paste0("c≈", round(c_setting, 1)),
    label      = paste0("w=(", w_label, "), ", c_label)
  )

n_agents_per_setting <- 5

simulation_plan <- param_grid |>
  uncount(n_agents_per_setting) |>
  mutate(agent_id = row_number())

sim_file <- here("advanced_cognitive_modeling/Assignment_4/simdata", "ch12_gcm_simulated_responses.csv")
sim_dir <- here("advanced_cognitive_modeling/Assignment_4/simdata")

if (!dir.exists(sim_dir)) dir.create(sim_dir)

if (regenerate_simulations || !file.exists(sim_file)) {
  plan(multisession, workers = max(1, availableCores() - 1))

  simulated_responses <- future_pmap_dfr(
    list(
      agent_id  = simulation_plan$agent_id,
      w_true    = simulation_plan$w_setting,
      c_true    = simulation_plan$c_setting,
      bias_true = simulation_plan$bias_setting
    ),
    simulate_gcm_agent,
    stimulus_info = stimulus_info,
    n_blocks      = n_blocks,
    subject_seed  = simulation_plan$agent_id,
    .options = furrr_options(seed = TRUE)
  )

  plan(sequential)
  write_csv(simulated_responses, sim_file)
  cat("Simulation complete. Results saved.\n")
} else {
  simulated_responses <- read_csv(sim_file, show_col_types = FALSE)
  cat("Loaded existing simulation results.\n")
}
```


```{r}
# plotting
simulated_responses_labeled <- simulated_responses |>
  mutate(
    w_label = paste0("w=(", 
                     round(w1_true, 2), ",", 
                     round(w2_true, 2), ",", 
                     round(w3_true, 2), ",", 
                     round(w4_true, 2), ",", 
                     round(w5_true, 2), ")"
                     ),
    c_label = paste0("c≈", round(c_true, 1)),
    # Order the c_label factor based on the numeric c_true values
    c_label = forcats::fct_reorder(c_label, c_true) 
  )

p_c_effect <- ggplot(
  simulated_responses_labeled,
  aes(x = trial_within_subject, y = performance,
      color = factor(round(c_true, 1)),
      group = interaction(agent_id, round(c_true, 1)))
) +
  stat_summary(fun = mean, geom = "line",
               aes(group = factor(round(c_true, 1))), linewidth = 1.2) +
  facet_wrap(~w_label) +
  scale_color_viridis_d(option = "plasma", name = "Sensitivity (c)") +
  labs(
    title    = "Effect of Sensitivity (c) on GCM Learning",
    subtitle = "Mean across simulated agents",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  theme(legend.position = "bottom")

p_w_effect <- ggplot(
  simulated_responses_labeled,
  aes(x = trial_within_subject, y = performance,
      color = w_label, group = interaction(agent_id, w_label))
) +
  stat_summary(fun = mean, geom = "line", aes(group = w_label), linewidth = 1.2) +
  facet_wrap(~c_label) +
  scale_color_brewer(palette = "Set1", name = "Attention Weights") +
  labs(
    title    = "Effect of Attention Weights (w) on GCM Learning",
    subtitle = "Mean across simulated agents",
    x = "Trial", y = "Cumulative Accuracy"
  ) +
  theme(legend.position = "bottom")

p_c_effect
p_w_effect
```
## Prior Predictive Check

*Note: * should we change the weight priors to actual random draws here??

```{r}
n_prior_samples <- 500

LOG_C_PRIOR_MEAN <- log(2)
LOG_C_PRIOR_SD <- 1

prior_draws <- tibble(
  sample_id = seq_len(n_prior_samples),
  log_c     = rnorm(n_prior_samples, LOG_C_PRIOR_MEAN, LOG_C_PRIOR_SD),
  w1        = 0.2,
  w2        = 0.2,
  w3        = 0.2,
  w4        = 0.2,
  w5        = 0.2,
  bias      = rbeta(n_prior_samples, 1, 1)
) |>
  mutate(c_true = exp(log_c))

ppc_schedule <- make_subject_schedule(stimulus_info, n_blocks, seed = 999)

prior_pred_curves <- pmap_dfr(
  list(
    sample_id = prior_draws$sample_id,
    log_c     = prior_draws$log_c,
    w1        = prior_draws$w1,
    w2        = prior_draws$w2,
    w3        = prior_draws$w3,
    w4        = prior_draws$w4,
    w5        = prior_draws$w5,
    bias      = prior_draws$bias
  ),
  function(sample_id, log_c, w1, w2, w3, w4, w5, bias) {
    c <- exp(log_c)
    sim <- gcm_simulate(
      w = c(w1, w2, w3, w4, w5), c = c, bias = bias,
      obs          = ppc_schedule |> dplyr::select(feature1, feature2, feature3, feature4, feature5),
      cat_feedback = ppc_schedule$category_feedback
    )
    tibble(
      sample_id = sample_id,
      trial     = seq_len(nrow(ppc_schedule)),
      correct   = as.integer(sim$sim_response == ppc_schedule$category_feedback)
    ) |>
      mutate(cum_acc = cumsum(correct) / row_number())
  }
)

prior_pred_summary <- prior_pred_curves |>
  group_by(trial) |>
  summarise(
    q05 = quantile(cum_acc, 0.05),
    q25 = quantile(cum_acc, 0.25),
    q50 = median(cum_acc),
    q75 = quantile(cum_acc, 0.75),
    q95 = quantile(cum_acc, 0.95),
    .groups = "drop"
  )

ggplot(prior_pred_summary, aes(x = trial)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#56B4E9", alpha = 0.20) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#56B4E9", alpha = 0.40) +
  geom_line(aes(y = q50), color = "#0072B2", linewidth = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "Prior Predictive Check: GCM Learning Curves",
    subtitle = "Ribbons: 50% and 90% prior predictive intervals; line: median\nPriors: log(c) ~ Normal(log 2, 1),  w ~ Dirichlet(1,1),  bias ~ Beta(1,1)",
    x = "Trial", y = "Cumulative Accuracy"
  )
```

## STAN model

```{r}
gcm_single_stan <- "
// Generalized Context Model — Single Subject (refactored architecture)

data {
  int<lower=1> N_total;
  int<lower=1> N_features;
  array[N_total] int<lower=0, upper=1> y;
  array[N_total, N_features] real obs;
  array[N_total] int<lower=0, upper=1> cat_feedback;

  vector[N_features] prior_w_alpha;
  real prior_log_c_mu;
  real<lower=0> prior_log_c_sigma;
  real<lower=0> prior_bias_alpha;
  real<lower=0> prior_bias_beta;
  int<lower=0, upper=1> run_diagnostics;
}

transformed data {
  array[N_total, N_total, N_features] real abs_diff;
  for (i in 1:N_total) {
    for (j in 1:N_total) {
      for (f in 1:N_features) {
        abs_diff[i, j, f] = abs(obs[i, f] - obs[j, f]);
      }
    }
  }
}

parameters {
  simplex[N_features] w;
  real log_c;
  real<lower=0, upper=1> bias;
}

transformed parameters {
  real<lower=0> c = exp(log_c);
  vector<lower=1e-9, upper=1-1e-9>[N_total] prob_cat1;

  {
    array[N_total] int memory_trial_idx;
    array[N_total] int memory_cat;
    int n_mem = 0;

    for (i in 1:N_total) {
      real p_i;
      int has_cat0 = 0;
      int has_cat1 = 0;

      for (k in 1:n_mem) {
        if (memory_cat[k] == 0) has_cat0 = 1;
        if (memory_cat[k] == 1) has_cat1 = 1;
      }

      if (n_mem == 0 || has_cat0 == 0 || has_cat1 == 0) {
        p_i = bias;
      } else {
        real s1 = 0; real s0 = 0;
        int n1 = 0;  int n0 = 0;
        for (e in 1:n_mem) {
          real d = 0;
          int past_i = memory_trial_idx[e];
          for (f in 1:N_features)
            d += w[f] * abs_diff[i, past_i, f];
          real sim = exp(-c * d);
          if (memory_cat[e] == 1) { s1 += sim; n1 += 1; }
          else                    { s0 += sim; n0 += 1; }
        }
        real m1 = s1 / n1;
        real m0 = s0 / n0;
        real num = bias * m1;
        real den = num + (1 - bias) * m0;
        p_i = (den > 1e-9) ? num / den : bias;
      }

      prob_cat1[i] = fmax(1e-9, fmin(1 - 1e-9, p_i));

      n_mem += 1;
      memory_trial_idx[n_mem] = i;
      memory_cat[n_mem] = cat_feedback[i];
    }
  }
}

model {
  target += dirichlet_lpdf(w    | prior_w_alpha);
  target += normal_lpdf(log_c   | prior_log_c_mu, prior_log_c_sigma);
  target += beta_lpdf(bias      | prior_bias_alpha,  prior_bias_beta);

  target += bernoulli_lpmf(y | prob_cat1);
}

generated quantities {
  vector[N_total] log_lik;
  real lprior;
  if (run_diagnostics) {
    for (i in 1:N_total)
      log_lik[i] = bernoulli_lpmf(y[i] | prob_cat1[i]);
  }
  lprior = dirichlet_lpdf(w | prior_w_alpha) +
           normal_lpdf(log_c | prior_log_c_mu, prior_log_c_sigma) +
           beta_lpdf(bias | prior_bias_alpha, prior_bias_beta);
}
"

stan_file_gcm_single <- "stan/ch12_gcm_single.stan"
write_stan_file(gcm_single_stan, dir = "stan/", basename = "ch12_gcm_single.stan")
```
```{r}
mod_gcm_single <- cmdstan_model(stan_file_gcm_single)
```

```{r}
# Select one simulated agent: w ≈ (0.9, 0.1), c ≈ 2
agent_to_fit <- simulated_responses |>
  filter(
    round(w4_true, 1) == 0.6,
    abs(log_c_true - log(2.0)) == min(abs(log_c_true - log(2.0)))
  ) |>
  filter(agent_id == min(agent_id)) |>
  slice(1:total_trials)

stopifnot(nrow(agent_to_fit) == total_trials)

gcm_data_single <- list(
  N_total           = nrow(agent_to_fit),
  N_features        = 5L,
  y                 = as.integer(agent_to_fit$sim_response),
  obs               = as.matrix(agent_to_fit[, c("feature1", "feature2", "feature3", "feature4", "feature5")]),
  cat_feedback      = as.integer(agent_to_fit$category_feedback),
  prior_w_alpha     = c(1, 1, 1, 1, 1),
  prior_log_c_mu    = LOG_C_PRIOR_MEAN,
  prior_log_c_sigma = LOG_C_PRIOR_SD,
  prior_bias_alpha  = 1,
  prior_bias_beta   = 1,
  run_diagnostics   = 1L
)

fit_filepath_single <- here("advanced_cognitive_modeling/Assignment_4/simmodels", "ch12_gcm_single_fit.rds")

if (regenerate_simulations || !file.exists(fit_filepath_single)) {
  pf_single <- tryCatch(
    mod_gcm_single$pathfinder(
      data      = gcm_data_single,
      seed      = 123,
      num_paths = 4,
      refresh   = 0
    ),
    error = function(e) {
      message("Pathfinder failed — falling back to default initialisation.")
      NULL
    }
  )

  fit_gcm_single <- mod_gcm_single$sample(
    data            = gcm_data_single,
    init            = pf_single,
    seed            = 123,
    chains          = 4,
    parallel_chains = min(4, availableCores()),
    iter_warmup     = 1000,
    iter_sampling   = 1500,
    refresh         = 300,
    adapt_delta     = 0.9
  )
  fit_gcm_single$save_object(fit_filepath_single)
  cat("Single-agent model fitted and saved.\n")
} else {
  fit_gcm_single <- readRDS(fit_filepath_single)
  cat("Loaded existing single-agent fit.\n")
}
```
## Diagnostics

```{r}
fit_gcm_single$summary()
```

*Note:* we cannot find diagnostic_summary_table function in the course notes. Maybe alter diagnostic function from ass3.

```{r}
bayesplot::mcmc_trace(
  fit_gcm_single$draws(c("log_c", "w[1]", "w[2]", "w[3]", "w[4]", "w[5]", "bias")),
  facet_args = list(ncol = 1)
)

bayesplot::mcmc_rank_overlay(
  fit_gcm_single$draws(c("log_c", "w[1]", "w[2]", "w[3]", "w[4]", "w[5]", "bias"))
)
```

## Posterior Predictive check

```{r}
if (!is.null(fit_gcm_single)) {
  prob_draws <- fit_gcm_single$draws("prob_cat1", format = "matrix")
  n_draws    <- nrow(prob_draws)
  n_trials   <- ncol(prob_draws)

  set.seed(2025)
  yrep_post <- matrix(rbinom(n_draws * n_trials, 1, as.vector(prob_draws)),
                      nrow = n_draws)

  obs_acc <- cummean(agent_to_fit$sim_response == agent_to_fit$category_feedback)
  rep_acc <- t(apply(yrep_post, 1, function(yr) {
    cummean(yr == agent_to_fit$category_feedback)
  }))

  ppc_summary <- tibble(
    trial = seq_len(n_trials),
    obs   = obs_acc,
    q05   = apply(rep_acc, 2, quantile, 0.05),
    q25   = apply(rep_acc, 2, quantile, 0.25),
    q50   = apply(rep_acc, 2, quantile, 0.50),
    q75   = apply(rep_acc, 2, quantile, 0.75),
    q95   = apply(rep_acc, 2, quantile, 0.95)
  )

  ggplot(ppc_summary, aes(x = trial)) +
    geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#56B4E9", alpha = 0.20) +
    geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#56B4E9", alpha = 0.40) +
    geom_line(aes(y = q50), color = "#0072B2", linewidth = 0.8) +
    geom_line(aes(y = obs), color = "#D55E00", linewidth = 1) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
      title    = "Posterior Predictive Check: Cumulative Accuracy",
      subtitle = "Blue ribbons: 50% and 90% posterior predictive intervals\nOrange line: observed cumulative accuracy",
      x = "Trial", y = "Cumulative Accuracy"
    )
}
```
## Prior-Posterior update check

*Note: * Wrap plots to grid. Fix W4 plot, estimate is way off.

```{r}
if (!is.null(fit_gcm_single)) {
  draws_df <- as_draws_df(
    fit_gcm_single$draws(variables = c("w[1]", "w[2]", "w[3]", "w[4]", "w[5]", "log_c", "c", "bias"))
  )

  true_w1    <- agent_to_fit$w1_true[1]
  true_w2    <- agent_to_fit$w2_true[1]
  true_w3    <- agent_to_fit$w3_true[1]
  true_w4    <- agent_to_fit$w4_true[1]
  true_w5    <- agent_to_fit$w5_true[1]
  true_log_c <- agent_to_fit$log_c_true[1]
  true_c     <- agent_to_fit$c_true[1]
  true_bias  <- agent_to_fit$bias_true[1]

  set.seed(2026)
  n_prior <- 4000
  prior_df <- tibble(
    `w[1]` = runif(n_prior),
    log_c  = rnorm(n_prior, LOG_C_PRIOR_MEAN, LOG_C_PRIOR_SD),
    c      = exp(log_c),
    bias   = rbeta(n_prior, 1, 1)
  )

  pp_panel <- function(prior_x, post_x, true_val, label) {
    tibble(
      x      = c(prior_x, post_x),
      source = c(rep("Prior", length(prior_x)),
                 rep("Posterior", length(post_x)))
    ) |>
      ggplot(aes(x = x, fill = source, color = source)) +
      geom_density(alpha = 0.4, linewidth = 0.5) +
      geom_vline(xintercept = true_val, color = "red",
                 linetype = "dashed", linewidth = 0.7) +
      scale_fill_manual(values = c(Prior = "grey60", Posterior = "#0072B2")) +
      scale_color_manual(values = c(Prior = "grey40", Posterior = "#0072B2")) +
      labs(title = label, x = NULL, y = "Density") +
      theme(legend.position = "bottom")
  }

  p1_1 <- pp_panel(prior_df$`w[1]`, draws_df$`w[1]`, true_w1, "w[1]")
  p1_2 <- pp_panel(prior_df$`w[1]`, draws_df$`w[2]`, true_w2, "w[2]")
  p1_3 <- pp_panel(prior_df$`w[1]`, draws_df$`w[3]`, true_w3, "w[3]")
  p1_4 <- pp_panel(prior_df$`w[1]`, draws_df$`w[4]`, true_w4, "w[4]")
  p1_5 <- pp_panel(prior_df$`w[1]`, draws_df$`w[5]`, true_w5, "w[5]")
  p2 <- pp_panel(prior_df$log_c,  draws_df$log_c,  true_log_c, "log(c) — computational scale")
  p3 <- pp_panel(prior_df$c,      draws_df$c,      true_c,     "c — natural scale")
  p4 <- pp_panel(prior_df$bias,   draws_df$bias,   true_bias,  "bias")

p1_1
p1_2
p1_3
p1_4
p1_5
p2
p3
p4

}
```
## Joint posterior geometry

*Note: * is this too wrong?

```{r}
#install.packages('hexbin')
library(hexbin)
if (!is.null(fit_gcm_single)) {
  bayesplot::mcmc_pairs(
    fit_gcm_single$draws(c("log_c", "w[1]", "w[2]", "w[3]", "w[4]", "w[5]", "bias")),
    diag_fun     = "dens",
    off_diag_fun = "hex"
  )
}
```
## LOO

*Note: * Cannot run, gives error about pathfinder. Unable to retrieve generated quantities. Maybe try pf_single instead of pf

```{r}
# ─────────────────────────────────────────────────────────────────────────────
# psis_lfo_gcm: PSIS-approximated leave-future-out CV for the GCM.
# Follows Bürkner, Gabry, & Vehtari (2020), Algorithm 1.
# ─────────────────────────────────────────────────────────────────────────────
psis_lfo_gcm <- function(stan_model,
                         full_data,
                         L = 16,
                         k_thresh = 0.7,
                         seed = 1,
                         verbose = FALSE) {

  fit_subset <- function(n_train, seed) {
    sub_data <- full_data
    sub_data$N_total      <- n_train
    sub_data$y            <- sub_data$y[1:n_train]
    sub_data$obs          <- sub_data$obs[1:n_train, , drop = FALSE]
    sub_data$cat_feedback <- sub_data$cat_feedback[1:n_train]

    pf  <- stan_model$pathfinder(data = sub_data, num_paths = 4, refresh = 0,
                                  seed = seed)
    fit <- stan_model$sample(
      data = sub_data, init = pf, seed = seed,
      chains = 2, parallel_chains = 2,
      iter_warmup = 600, iter_sampling = 800,
      refresh = 0, adapt_delta = 0.9
    )
    fit
  }

  predict_future <- function(fit, full_data) {
    full_n   <- length(full_data$y)
    ext_data <- full_data
    ext_data$N_total <- full_n

    gq_fit <- stan_model$generate_quantities(
      fitted_params   = fit,
      data            = ext_data,
      seed            = seed,
      parallel_chains = 2
    )
    gq_fit$draws("log_lik", format = "matrix")
  }

  T_total      <- length(full_data$y)
  elpd_lfo     <- numeric(T_total)
  elpd_lfo[]   <- NA_real_
  refits       <- integer(0)
  k_history    <- numeric(0)

  fit_curr     <- fit_subset(L, seed = seed)
  ll_full_curr <- predict_future(fit_curr, full_data)

  refits       <- c(refits, L)
  last_fit_n   <- L
  n_draws      <- nrow(ll_full_curr)

  ll_train_curr <- rowSums(ll_full_curr[, 1:L, drop = FALSE])

  for (t in (L + 1):T_total) {
    if (t - 1 == last_fit_n) {
      k_t          <- 0
      k_history[t] <- k_t
      lw           <- rep(-log(n_draws), n_draws)
    } else {
      ll_train_target <- rowSums(ll_full_curr[, 1:(t - 1), drop = FALSE])
      log_w           <- ll_train_target - ll_train_curr

      psis_obj     <- loo::psis(log_w, r_eff = NA)
      k_t          <- loo::pareto_k_values(psis_obj)
      k_history[t] <- k_t

      if (k_t > k_thresh) {
        if (verbose) cat("Refit at t =", t, "(k =", round(k_t, 3), ")\n")
        fit_curr      <- fit_subset(t - 1, seed = seed + t)
        ll_full_curr  <- predict_future(fit_curr, full_data)
        ll_train_curr <- rowSums(ll_full_curr[, 1:(t - 1), drop = FALSE])
        last_fit_n    <- t - 1
        refits        <- c(refits, last_fit_n)
        lw            <- rep(-log(n_draws), n_draws)
      } else {
        lw <- as.numeric(weights(psis_obj))
      }
    }

    lp          <- ll_full_curr[, t]
    elpd_lfo[t] <- matrixStats::logSumExp(lw + lp)
  }

  list(
    elpd_lfo       = sum(elpd_lfo, na.rm = TRUE),
    elpd_pointwise = elpd_lfo,
    refits         = refits,
    k_history      = k_history,
    L              = L
  )
}
```


```{r}
lfo_filepath <- here("simdata", "ch12_lfo_single.rds")
if (regenerate_simulations || !file.exists(lfo_filepath)) {
  lfo_single <- psis_lfo_gcm(
    stan_model = mod_gcm_single,
    full_data  = gcm_data_single,
    L          = 16,
    k_thresh   = 0.7,
    seed       = 11,
    verbose    = TRUE
  )
  saveRDS(lfo_single, lfo_filepath)
} else {
  lfo_single <- readRDS(lfo_filepath)
}

cat("LFO-CV ELPD (single subject):", round(lfo_single$elpd_lfo, 2), "\n")

cat("Number of refits:",             length(lfo_single$refits),     "\n")

cat("Refit triggered at trials:",    lfo_single$refits,             "\n")
```
```{r}
if (!is.null(fit_gcm_single)) {
  loo_pw  <- loo_gcm_single$pointwise[, "elpd_loo"]
  comp_df <- tibble(
    trial    = seq_along(loo_pw),
    elpd_loo = loo_pw,
    elpd_lfo = lfo_single$elpd_pointwise
  ) |>
    pivot_longer(starts_with("elpd_"), names_to = "method", values_to = "elpd")

  ggplot(comp_df, aes(x = trial, y = elpd, color = method)) +
    geom_line(alpha = 0.7) +
    geom_point(alpha = 0.5) +
    scale_color_manual(values = c(elpd_loo = "#0072B2", elpd_lfo = "#D55E00"),
                       labels = c("PSIS-LFO (correct)",
                                  "PSIS-LOO (invalid for sequential)")) +
    labs(
      title    = "Per-trial ELPD: LOO vs LFO-CV",
      subtitle = "Differences are largest where the path-dependence assumption fails most",
      x = "Trial", y = "Pointwise ELPD"
    ) +
    theme(legend.position = "bottom")
}
```

## Fitting model to empirical data

*Note: * We are supposed to fit the model by participant, how do we do that?

```{r}
# Select one simulated agent: w ≈ (0.9, 0.1), c ≈ 2
empirical_agent_to_fit <- df |>
  filter(
    subject == 1
  )

empirical_gcm_data_single <- list(
  N_total           = nrow(empirical_agent_to_fit),
  N_features        = 5L,
  y                 = as.integer(empirical_agent_to_fit$response_danger),
  obs               = as.matrix(empirical_agent_to_fit[, c("feature1", "feature2", "feature3", "feature4", "feature5")]),
  cat_feedback      = as.integer(empirical_agent_to_fit$dangerous),
  prior_w_alpha     = c(1, 1, 1, 1, 1),
  prior_log_c_mu    = LOG_C_PRIOR_MEAN,
  prior_log_c_sigma = LOG_C_PRIOR_SD,
  prior_bias_alpha  = 1,
  prior_bias_beta   = 1,
  run_diagnostics   = 1L
)

empirical_fit_filepath_single <- here("advanced_cognitive_modeling/Assignment_4/simmodels", "empirical_gcm_single_fit.rds")

if (regenerate_simulations || !file.exists(empirical_fit_filepath_single)) {
  empirical_pf_single <- tryCatch(
    mod_gcm_single$pathfinder(
      data      = empirical_gcm_data_single,
      seed      = 123,
      num_paths = 4,
      refresh   = 0
    ),
    error = function(e) {
      message("Pathfinder failed — falling back to default initialisation.")
      NULL
    }
  )

  empirical_fit_gcm_single <- mod_gcm_single$sample(
    data            = empirical_gcm_data_single,
    init            = empirical_pf_single,
    seed            = 123,
    chains          = 4,
    parallel_chains = min(4, availableCores()),
    iter_warmup     = 1000,
    iter_sampling   = 1500,
    refresh         = 300,
    adapt_delta     = 0.9
  )
  empirical_fit_gcm_single$save_object(empirical_fit_filepath_single)
  cat("Single-agent model fitted and saved.\n")
} else {
  empirical_fit_gcm_single <- readRDS(empirical_fit_filepath_single)
  cat("Loaded existing single-agent fit.\n")
}
```
## Diagnostics

```{r}
empirical_fit_gcm_single$summary()
```

```{r}
bayesplot::mcmc_trace(
  empirical_fit_gcm_single$draws(c("log_c", "w[1]", "w[2]", "w[3]", "w[4]", "w[5]", "bias")),
  facet_args = list(ncol = 1)
)

bayesplot::mcmc_rank_overlay(
  empirical_fit_gcm_single$draws(c("log_c", "w[1]", "w[2]", "w[3]", "w[4]", "w[5]", "bias"))
)
```

## Posterior Predictive check

```{r}
if (!is.null(empirical_fit_gcm_single)) {
  prob_draws <- empirical_fit_gcm_single$draws("prob_cat1", format = "matrix")
  n_draws    <- nrow(prob_draws)
  n_trials   <- ncol(prob_draws)

  set.seed(2025)
  yrep_post <- matrix(rbinom(n_draws * n_trials, 1, as.vector(prob_draws)),
                      nrow = n_draws)

  obs_acc <- cummean(empirical_agent_to_fit$response_danger == empirical_agent_to_fit$dangerous)
  rep_acc <- t(apply(yrep_post, 1, function(yr) {
    cummean(yr == empirical_agent_to_fit$dangerous)
  }))

  ppc_summary <- tibble(
    trial = seq_len(n_trials),
    obs   = obs_acc,
    q05   = apply(rep_acc, 2, quantile, 0.05),
    q25   = apply(rep_acc, 2, quantile, 0.25),
    q50   = apply(rep_acc, 2, quantile, 0.50),
    q75   = apply(rep_acc, 2, quantile, 0.75),
    q95   = apply(rep_acc, 2, quantile, 0.95)
  )

  ggplot(ppc_summary, aes(x = trial)) +
    geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#56B4E9", alpha = 0.20) +
    geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#56B4E9", alpha = 0.40) +
    geom_line(aes(y = q50), color = "#0072B2", linewidth = 0.8) +
    geom_line(aes(y = obs), color = "#D55E00", linewidth = 1) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
      title    = "Empirical Posterior Predictive Check: Cumulative Accuracy",
      subtitle = "Blue ribbons: 50% and 90% posterior predictive intervals\nOrange line: observed cumulative accuracy",
      x = "Trial", y = "Cumulative Accuracy"
    )
}
```
## Prior-posterior update check

*Note: * Vi er nået hertil.

```{r}
if (!is.null(empirical_fit_gcm_single)) {
  draws_df <- as_draws_df(
    empirical_fit_gcm_single$draws(variables = c("w[1]", "w[2]", "w[3]", "w[4]", "w[5]", "log_c", "c", "bias"))
  )

  true_w1    <- agent_to_fit$w1_true[1]
  true_w2    <- agent_to_fit$w2_true[1]
  true_w3    <- agent_to_fit$w3_true[1]
  true_w4    <- agent_to_fit$w4_true[1]
  true_w5    <- agent_to_fit$w5_true[1]
  true_log_c <- agent_to_fit$log_c_true[1]
  true_c     <- agent_to_fit$c_true[1]
  true_bias  <- agent_to_fit$bias_true[1]

  set.seed(2026)
  n_prior <- 4000
  prior_df <- tibble(
    `w[1]` = runif(n_prior),
    log_c  = rnorm(n_prior, LOG_C_PRIOR_MEAN, LOG_C_PRIOR_SD),
    c      = exp(log_c),
    bias   = rbeta(n_prior, 1, 1)
  )

  pp_panel <- function(prior_x, post_x, true_val, label) {
    tibble(
      x      = c(prior_x, post_x),
      source = c(rep("Prior", length(prior_x)),
                 rep("Posterior", length(post_x)))
    ) |>
      ggplot(aes(x = x, fill = source, color = source)) +
      geom_density(alpha = 0.4, linewidth = 0.5) +
      geom_vline(xintercept = true_val, color = "red",
                 linetype = "dashed", linewidth = 0.7) +
      scale_fill_manual(values = c(Prior = "grey60", Posterior = "#0072B2")) +
      scale_color_manual(values = c(Prior = "grey40", Posterior = "#0072B2")) +
      labs(title = label, x = NULL, y = "Density") +
      theme(legend.position = "bottom")
  }

  p1_1 <- pp_panel(prior_df$`w[1]`, draws_df$`w[1]`, true_w1, "w[1]")
  p1_2 <- pp_panel(prior_df$`w[1]`, draws_df$`w[2]`, true_w2, "w[2]")
  p1_3 <- pp_panel(prior_df$`w[1]`, draws_df$`w[3]`, true_w3, "w[3]")
  p1_4 <- pp_panel(prior_df$`w[1]`, draws_df$`w[4]`, true_w4, "w[4]")
  p1_5 <- pp_panel(prior_df$`w[1]`, draws_df$`w[5]`, true_w5, "w[5]")
  p2 <- pp_panel(prior_df$log_c,  draws_df$log_c,  true_log_c, "log(c) — computational scale")
  p3 <- pp_panel(prior_df$c,      draws_df$c,      true_c,     "c — natural scale")
  p4 <- pp_panel(prior_df$bias,   draws_df$bias,   true_bias,  "bias")

p1_1
p1_2
p1_3
p1_4
p1_5
p2
p3
p4

}
```





