---
title: 'Assignment 2: STAN'
author: "Study Group 6"
date: "2026-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Loading required packages
pacman::p_load(
  tidyverse,    # Data manipulation and plotting
  here,         # Robust file paths
  cmdstanr,     # Stan interface
  posterior,    # Posterior analysis
  bayesplot,    # Plotting MCMC
  furrr,        # Parallel processing
  future        # Future backend
)
```

Simulation code from the google docs

```{r reinforcement agent}
set.seed(123)

trials <- 20  

# random agent director
opponents_choices = rbinom(trials, 1, 0.8)

# parameters to be recovered
alpha <- 0.8 # learning rate
tau <- 1 # inverse temperature

# prepare variables
V <- numeric(trials) # value
p <- numeric(trials) # probability of choosing right
choices <- integer(trials) # guesser's choices

# first trial
V[1] = 0.5 # uninformed starting value
p[1] <- 1 / (1 + exp(-tau * V[1])) # softmax - tehcnically obsolete with a starting v of 0.5 and binary choice
choices[1] <- rbinom(1, 1, p[1]) # making the first choice

for (t in 2:trials) {
  
  # belief update (Rescorla-Wagner)
  V[t] = V[t-1] + alpha * (opponents_choices[t-1] - V[t-1])

  # prob of choosing right (softmax)
  p[t] = 1 / (1 + exp(-tau * V[t]))

  # make choice
	choices[t] = rbinom(1, 1, p[t])
}

# saving to df
df <- data.frame(
  trial = 1:trials, 
  value = V,
  prob = p,
  opponents_choice = factor(opponents_choices,
                  levels = c(0, 1),
                  labels = c("left", "right")), # rename 1 to right, and 0 to left 
  choice = factor(choices,
                  levels = c(0, 1),
                  labels = c("left", "right"))
  )

```

```{r}
# adding some extra quantities to the tibble
df$choice_numeric <- ifelse(df$choice == "right", 1, 0)
df$opponents_choice_numeric <- ifelse(df$opponents_choice == "right", 1, 0)
df$reward_right <- ifelse(df$opponents_choice == "right", 1, 0)
```


Plotting similarly to in assignment 1

```{r}


choice_distribution <- ggplot(df, aes(x = trial)) +
  geom_line(aes(y = value, color = "Expected value of right"), linewidth = 0.8) +  # draw line FIRST (bottom layer)
  geom_point(aes(y = opponents_choice_numeric, color = "Biased opponent"), alpha = 0.6) +
  geom_point(aes(y = choice_numeric, color = "RL agent"), alpha = 0.6) +
  scale_y_continuous(
    limits = c(-0.1, 1.1),
    breaks = c(0, 1),
    labels = c("Left (0)", "Right (1)")
  ) +
  labs(title = "RL agent vs. Biased Random", y = "Choice") +
  theme_classic()

choice_distribution




```
SO the agent changes its expect value very quickly
could also plot the relationship between V and p, which is determined by tau, and the relationship between correctness and V


### writing the stan-code
In the below i took from course notes 5.4 the structure of the r-to-stan pipeline and added own stan code based on whats in the docs, updated to hopefulle work

```{r}
stan_model_code <- "
// 1. Data Block: Declares the data Stan expects from R
data {
  int<lower=1> n;       		   // Number of trials (must be at least 1)
  array[n] int<lower=0, upper=1> h; // Array 'h' of length 'n' containing choices (0 or 1)
  array[n] int<lower=0, upper=1> h_opponent; // Array 'h' of length 'n' containing opponents choices (0 or 1)
  array[n] int<lower=0, upper=1> feedback; // Array 'h' of length 'n' containing opponents choices (0 or 1)
}

parameters {
real <lower=0, upper = 1> alpha; 	    	  //not a vector → Learning rate is fixed (1 no.)
real <lower=0, upper = 1> tau;		  //not a vector → no change in outcome after 20
}

transformed parameters {
real <lower=0, upper=20> tau_scaled = tau*20;   //rescale tau to 0-20
}


model {
  array[n] real V;

  // Prior: Our belief about theta *before* seeing the data.
  // 'target +=' adds the log-probability density to the overall model log-probability.

  target += beta_lpdf(alpha | 1, 1); // lpdf = log probability density function
  target += beta_lpdf(tau | 2, 8);


  // Likelihood: How the data 'h' depend on the parameter 'theta'.
  // The model assesses how likely the observed sequence 'h' is given a value of 'theta'.

  V[1] = 0.5;
  target += bernoulli_lpmf(h[1]| 0.5); // lpmf = log probability mass function (for discrete data)

  for (t in 2:n){
    V[t]=V[t-1]+alpha*(feedback[t-1]-V[t-1]); // reward and expected value are both for picking right
    real p=1/(1+exp(-tau_scaled*V[t]));

  target += bernoulli_lpmf(h[t]| p); // lpmf = log probability mass function (for discrete data)
  }
}
"
# Define relative path for Stan model file
stan_model_dir <- "advanced_cognitive_modeling/Assignment_2/stan"
stan_file_RL <- here(stan_model_dir, "RL_model.stan")

# Create directory if needed
if (!dir.exists(stan_model_dir)) dir.create(stan_model_dir)

# Write the Stan code to the file
writeLines(stan_model_code, stan_file_RL)

cat("Stan model written to:", stan_file_RL, "\n")
```
Had some trouble with the file- paths but it seems to work now

```{r}
# Stan requires data in a list format
data_for_stan <- list(
  n = trials,     # Number of trials (integer)
  h = df$choice_numeric,
  h_opponent = df$opponents_choice_numeric,
  feedback = df$reward_right
)

str(data_for_stan) # Show the structure of the list

```




## fitting

```{r}
## --- Compile and Fit ---
# Specify model file path
stan_file_RL <- here(stan_model_dir, "RL_model.stan")

# Define path for saving fitted model object
stan_results_dir <- "simmodels"
model_file_RL <- here(stan_results_dir, "RL_model.rds")

if (!dir.exists(stan_results_dir)) dir.create(stan_results_dir)

if (!file.exists(model_file_RL)) {
  cat("Compiling and fitting Stan model (RL)...\n")
  # Compile the Stan model (only needs to be done once unless code changes)
  mod_RL <- cmdstan_model(stan_file_RL,
                                 cpp_options = list(stan_threads = FALSE), # Enable threading
                                 stanc_options = list("O1")) # Basic optimization

  # Sample from the posterior distribution using MCMC
  fit_RL <- mod_RL$sample(
    data = data_for_stan,          # The data list we prepared
    seed = 123,                    # For reproducible MCMC sampling
    chains = 4,                    # Number of parallel Markov chains (recommend 4)
    parallel_chains = 4, # Run chains in parallel
    #threads_per_chain = 1,         # For within-chain parallelization (usually 1 is fine)
    iter_warmup = 1000,            # Number of warmup iterations per chain (discarded)
    iter_sampling = 2000,          # Number of sampling iterations per chain (kept)
    refresh = 0,                   # How often to print progress updates. 0 means never since the model is so fast
    max_treedepth = 10,            # Controls complexity of MCMC steps (adjust if needed)
    adapt_delta = 0.8              # Target acceptance rate (adjust if divergences occur)
  )

  # Save the fitted model object
  fit_RL$save_object(file = model_file_bernoulli)
  cat("Model fit completed and saved to:", model_file_RL, "\n")
} else {
  # Load existing fitted model object
  fit_RL <- readRDS(model_file_RL)
  cat("Loaded existing model fit from:", model_file_RL, "\n")
}
```





