---
title: 'Assignment 2: STAN'
author: "Study Group 6"
date: "2026-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Loading required packages
pacman::p_load(
  tidyverse,    # Data manipulation and plotting
  here,         # Robust file paths
  cmdstanr,     # Stan interface
  posterior,    # Posterior analysis
  bayesplot,    # Plotting MCMC
  furrr,        # Parallel processing
  future        # Future backend
)
```

```{r}
# adding a regenerate simulations variable to use riccardos structure - this should be changed  to true every time the stan code is changed or somethig else that require things to be simulated again BY SETTING THIS TO TRUE EVERYTHING SHOULD RUN WITHOUT HAVING TO DELETE THE MODEL FILES IN THE FOLDERS

regenerate_simulations <- TRUE
```

```{r reinforcement agent}
set.seed(123)

trials <- 500  

# random biased agent director
opponents_choices = rbinom(trials, 1, 0.8)

# parameters to be recovered
alpha <- 0.8 # learning rate
tau <- 1 # inverse temperatur

simulate_rl_data <- function(trials, alpha, tau, opponents_choices){

# prepare variables
V <- numeric(trials) # value
p <- numeric(trials) # probability of choosing right
choices <- integer(trials) # guesser's choices

# first trial
V[1] = 0.5 # uninformed starting value
p[1] <- 1 / (1 + exp(-tau * V[1])) # softmax - technically obsolete with a starting v of 0.5 and binary choice
choices[1] <- rbinom(1, 1, p[1]) # making the first choice

for (t in 2:trials) {
  
  # belief update (Rescorla-Wagner)
  V[t] = V[t-1] + alpha * (opponents_choices[t-1] - V[t-1])

  # prob of choosing right (softmax)
  p[t] = 1 / (1 + exp(-tau * V[t]))

  # make choice
	choices[t] = rbinom(1, 1, p[t])
	
}




# saving to df
df <- data.frame(
  trial = 1:trials, 
  value = V,
  prob = p,
  opponents_choice = factor(opponents_choices,
                  levels = c(0, 1),
                  labels = c("left", "right")), # rename 1 to right, and 0 to left 
  choice = factor(choices,
                  levels = c(0, 1),
                  labels = c("left", "right"))
  )

# adding some extra quantities to the tibble
df$opponents_choice_numeric <- ifelse(df$opponents_choice == "right", 1, 0)
df$reward_right <- ifelse(df$opponents_choice == "right", 1, 0)
df$choice_numeric <- ifelse(df$choice == "right", 1, 0)

return(df)

}

df <- simulate_rl_data(trials=trials, alpha=alpha, tau=tau, opponents_choices = opponents_choices )

```


Plotting similarly to in assignment 1

```{r}
choice_distribution <- ggplot(df, aes(x = trial)) +
  geom_line(aes(y = value, color = "Expected value of right"), linewidth = 0.8) +  # draw line FIRST (bottom layer)
  geom_point(aes(y = opponents_choice_numeric, color = "Biased opponent"), alpha = 0.6, size = 2, position = position_jitter(width = 0.1, height = 0.05)) +
  geom_point(aes(y = choice_numeric, color = "RL agent"), alpha = 0.6, size = 2, position = position_jitter(width = 0.1, height = 0.05)) +
  scale_y_continuous(
    limits = c(-0.1, 1.1),
    breaks = c(0, 1),
    labels = c("Left (0)", "Right (1)")
  ) +
  labs(title = "RL agent vs. Biased Random", y = "Choice") +
  theme_classic()

choice_distribution

```
# Nadia: hello guys added this to help justify our priors
```{r}
library(tidyverse)

# x-grid 
alpha_grid <- seq(0, 1, length.out = 1000)

# densities for Beta priors
beta_df <- bind_rows(
  tibble(
    alpha = alpha_grid,
    density = dbeta(alpha_grid, 1, 4),
    prior = "Beta(1,4)"
  ),
  tibble(
    alpha = alpha_grid,
    density = dbeta(alpha_grid, 2, 2),
    prior = "Beta(2,2)"
  ),
  tibble(
    alpha = alpha_grid,
    density = dbeta(alpha_grid, 3, 6),
    prior = "Beta(3,6)"
  ),
  tibble(
    alpha = alpha_grid,
    density = dbeta(alpha_grid, 4, 6),
    prior = "Beta(4,6)"
  )
)

# set legend order
beta_df$prior <- factor(
  beta_df$prior,
  levels = c(
    "Beta(1,4)",
    "Beta(2,2)",
    "Beta(3,6)",
    "Beta(4,6)"
  )
)

# plot
ggplot(beta_df, aes(x = alpha, y = density, color = prior)) +
  geom_line(linewidth = 1.5) +
  labs(
    title = "Beta distribution shapes for alpha prior",
    x = "Alpha (learning rate)",
    y = "Density",
    color = "Prior"
  ) +
  theme_minimal(base_size = 18)
```

# plot for report
```{r}
library(tidyverse)

# Generate x values
x <- seq(0.001, 0.999, length.out = 1000)

# Compute densities
prior_df <- bind_rows(
  tibble(
    x       = x,
    density = dbeta(x, 2, 2),
    prior   = "α ~ Beta(2,2) — learning rate"
  ),
  tibble(
    x       = x,
    density = dbeta(x, 2, 8),
    prior   = "τ ~ Beta(2,8) — inverse temperature"
  )
)

# Plot
ggplot(prior_df, aes(x = x, y = density, color = prior, fill = prior)) +
  geom_line(linewidth = 1.2) +
  geom_area(alpha = 0.15, position = "identity") +
  scale_color_manual(values = c(
    "α ~ Beta(2,2) — learning rate"              = "#378ADD",
    "τ ~ Beta(2,8) — inverse temperature"  = "#1D9E75"
  )) +
  scale_fill_manual(values = c(
    "α ~ Beta(2,2) — learning rate"              = "#378ADD",
    "τ ~ Beta(2,8) — inverse temperature"  = "#1D9E75"
  )) +
  labs(
    title   = "Prior distributions",
    x       = "Parameter value (0 – 1)",
    y       = "Density",
    color   = NULL,
    fill    = NULL
  ) +
  theme_ipsum() +
  theme(legend.position = "bottom")
```



So the agent changes its expected value very quickly.

Could also plot the relationship between V and p, which is determined by tau, and the relationship between correctness and V


### writing the stan-code
In the below i took from course notes 5.4 the structure of the r-to-stan pipeline and added own stan code based on whats in the docs, updated to hopefulle work

```{r}
stan_model_code <- "

// 1. Data Block: Declares the data Stan expects from R
data {
  int<lower=1> n;       		   // Number of trials (must be at least 1)
  array[n] int<lower=0, upper=1> h; // Array 'h' of length 'n' containing choices (0 or 1)
  array[n] int<lower=0, upper=1> h_opponent; // Array 'h' of length 'n' containing opponents choices (0 or 1)
  array[n] int<lower=0, upper=1> feedback; // Array 'h' of length 'n' containing opponents choices (0 or 1)
  real <lower = 0> prior_alpha_a; // added prior variable for alpha - it might be good to add for tau as well
  real <lower = 0> prior_alpha_b;
  
}

parameters {
real <lower=0, upper = 1> alpha; 	    	  //not a vector → Learning rate is fixed (1 no.)
real <lower=0, upper = 1> tau;		  //not a vector → no change in outcome after 20
}

transformed parameters {
real <lower=0, upper=20> tau_scaled = tau*20;   //rescale tau to 0-20
}


model {
  array[n] real V;

  // Prior: Our belief about alpha and tau *before* seeing the data.
  // 'target +=' adds the log-probability density to the overall model log-probability.

  target += beta_lpdf(alpha | prior_alpha_a, prior_alpha_b); // lpdf = log probability density function
  target += beta_lpdf(tau | 2, 8);


  // Likelihood: How the data 'h' depend on the parameter 'theta'.
  // The model assesses how likely the observed sequence 'h' is given a value of 'theta'.

  V[1] = 0.5;
  target += bernoulli_lpmf(h[1]| 0.5); // lpmf = log probability mass function (for discrete data)

  for (t in 2:n){
    V[t]=V[t-1]+alpha*(feedback[t-1]-V[t-1]); // reward and expected value are both for picking right
    real p=1/(1+exp(-tau_scaled*V[t]));

  target += bernoulli_lpmf(h[t]| p); // lpmf = log probability mass function (for discrete data)
  }
  
}

generated quantities {
  real<lower=0, upper=1> alpha_prior;
  real<lower=0, upper=1> tau_prior;
  real<lower=0, upper=20> tau_prior_scaled;

  alpha_prior = beta_rng(prior_alpha_a, prior_alpha_b); // Not uniform - we expect most people to have a moderate learning rate
  tau_prior = beta_rng(2, 8);
  tau_prior_scaled = tau_prior * 20;
  
}
"


# Define relative path for Stan model file
#stan_model_dir <- "/Users/nadiahajighassem/Documents/GitHUb/CogSci_MSc2/advanced_cognitive_modeling/Assignment_2/stan"
stan_model_dir <- "advanced_cognitive_modeling/Assignment_2/stan"
stan_file_RL <- here(stan_model_dir, "RL_model.stan")

# Create directory if needed
if (!dir.exists(stan_model_dir)) dir.create(stan_model_dir)

# Write the Stan code to the file
writeLines(stan_model_code, stan_file_RL)

cat("Stan model written to:", stan_file_RL, "\n")
```
Had some trouble with the file- paths but it seems to work now

```{r}
# Stan requires data in a list format
data_for_stan <- list(
  n = trials,     # Number of trials (integer)
  h = df$choice_numeric,
  h_opponent = df$opponents_choice_numeric,
  feedback = df$reward_right,
  prior_alpha_a = 2,
  prior_alpha_b = 2
)

str(data_for_stan) # Show the structure of the list

```
## fitting

```{r}
## --- Compile and Fit ---
# Specify model file path
stan_file_RL <- here(stan_model_dir, "RL_model.stan")

# Define path for saving fitted model object
stan_results_dir <- "~/Documents/GitHub/CogSci_MSc2/advanced_cognitive_modeling/Assignment_2/simmodels"
model_file_RL <- file.path(stan_results_dir, "RL_model1.rds")


if (!dir.exists(stan_results_dir)) dir.create(stan_results_dir)

if (regenerate_simulations | !file.exists(model_file_RL)) {
  cat("Compiling and fitting Stan model (RL)...\n")
  # Compile the Stan model (only needs to be done once unless code changes)
  mod_RL <- cmdstan_model(stan_file_RL,
                                 cpp_options = list(stan_threads = FALSE), # Enable threading
                                 stanc_options = list("O1")) # Basic optimization

  # Sample from the posterior distribution using MCMC
  fit_RL <- mod_RL$sample(
    data = data_for_stan,          # The data list we prepared
    seed = 123,                    # For reproducible MCMC sampling
    chains = 4,                    # Number of parallel Markov chains (recommend 4)
    parallel_chains = 4, # Run chains in parallel
    #threads_per_chain = 1,         # For within-chain parallelization (usually 1 is fine)
    iter_warmup = 1000,            # Number of warmup iterations per chain (discarded)
    iter_sampling = 2000,          # Number of sampling iterations per chain (kept)
    refresh = 0,                   # How often to print progress updates. 0 means never since the model is so fast
    max_treedepth = 10,            # Controls complexity of MCMC steps (adjust if needed)
    adapt_delta = 0.8              # Target acceptance rate (adjust if divergences occur)
  )

  # Save the fitted model object
  fit_RL$save_object(file = model_file_RL)
  cat("Model fit completed and saved to:", model_file_RL, "\n")
} else {
  # Load existing fitted model object
  fit_RL <- readRDS(model_file_RL)
  cat("Loaded existing model fit from:", model_file_RL, "\n")
}
```

```{r}
#cmdstanr::rebuild_cmdstan()
```


## Quality Assessment 

```{r}
# 1. Check for Divergent Transitions
diagnostics <- fit_RL$diagnostic_summary()
cat("Divergent transitions per chain:", paste(diagnostics$num_divergent, collapse = ", "), "\n")
```

```{r}
# 2. Check R-hat and ESS metrics
fit_summary <- fit_RL$summary(variables = c("alpha", "tau", "tau_scaled"))

fit_summary[, c("variable", "mean", "rhat", "ess_bulk", "ess_tail")]
```
```{r}
# 3. Visual Checks (Trace and Rank Plots) - For tau

draws_df <- as_draws_df(fit_RL$draws())

# A healthy trace plot is well-mixed with no wandering
p_trace <- mcmc_trace(draws_df, pars = "tau_scaled") + 
  theme_classic() + 
  ggtitle("Trace Plot: Well-mixed chains")

# A healthy rank histogram has uniform bars across all chains
p_rank <- mcmc_rank_hist(draws_df, pars = "tau_scaled") + 
  theme_classic() + 
  ggtitle("Rank Histogram: Chains explored the same space")

p_trace 
p_rank
```

```{r}
# For alpha

# A healthy trace plot is well-mixed with no wandering
p_trace <- mcmc_trace(draws_df, pars = "alpha") + 
  theme_classic() + 
  ggtitle("Trace Plot: Well-mixed chains")

# A healthy rank histogram has uniform bars across all chains
p_rank <- mcmc_rank_hist(draws_df, pars = "alpha") + 
  theme_classic() + 
  ggtitle("Rank Histogram: Chains explored the same space")

p_trace 
p_rank
```



```{r}
# 4. Look at the Posterior Estimate (Alpha)
ggplot(draws_df) +
  geom_density(aes(alpha, fill = "Posterior"), alpha = 0.6) +
  geom_density(aes(alpha_prior, fill = "Prior"), alpha = 0.6) +
  geom_vline(xintercept = 0.8, linetype = "dashed", color = "black", linewidth = 1.2) +
  scale_fill_manual(values = c("Posterior" = "blue", "Prior" = "red")) +
  labs(
    title = "Prior-Posterior Update for alpha",
    x = "Learning Rate (alpha)",
    y = "Density",
    fill = "Distribution"
  ) +
  theme_classic()

#Posterior Estimate (Tau)
ggplot(draws_df) +
  geom_density(aes(x = tau_scaled, fill = "Posterior"), alpha = 0.6) +
  geom_density(aes(x = tau_prior_scaled, fill = "Prior"), alpha = 0.6) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black", linewidth = 1.2) +
  scale_fill_manual(values = c("Posterior" = "blue", "Prior" = "red")) +
  labs(
    title = "Prior-Posterior Update for tau_scaled",
    x = "Temperature",
    y = "Density",
    fill = "Distribution"
  ) +
  theme_classic()
                    
```
So it seems the model has learned the tau with great precision but it is learning less about the alfa

## Carls old code for prior predictive check
```{r}
# This codes was the first attempt at prior predictive checks - it doesnt work
# Extract the summary statistic calculated from prior predictive simulations
# 'prior_rep_sum' was calculated within the generated quantities block

# Ensure we have the draws as a dataframe
#observed_sum <- sum(data_for_stan$h) # Total number of 'right' choices in the actual observed data

#ppc_prior_ggplot <- ggplot(draws_df, aes(x = prior_rep_sum)) +
  # Draw the histogram of the model's prior predictions (light red)
#  geom_histogram(fill = "#ffbaba", color = "#ff5252", bins = 30, alpha = 0.8) +
  
  # Add the vertical line for the actual observed data (dark red)
 # geom_vline(xintercept = observed_sum, color = "darkred", linewidth = 1.2) +
  
  # Labels and theming
  #ggtitle("Prior Predictive Check: Total 'Right' Choices") +
  #xlab(paste("Total Number of 'Right' Choices (out of", trials, "trials)")) +
  #ylab("Frequency") +
  #theme_classic()

#ppc_prior_ggplot
```
--------------------------------------

## NADIA Prior predictive check
```{r}
stan_model_code <- "
data {
  int<lower=1> n;
  array[n] int<lower=0, upper=1> h;
  array[n] int<lower=0, upper=1> h_opponent;
  array[n] int<lower=0, upper=1> feedback;
  real<lower=0> prior_alpha_a;
  real<lower=0> prior_alpha_b;
}

parameters {
  real<lower=0, upper=1> alpha;
  real<lower=0, upper=1> tau;
}

transformed parameters {
  real<lower=0, upper=20> tau_scaled = tau * 20;
}

model {
  array[n] real V;

  target += beta_lpdf(alpha | prior_alpha_a, prior_alpha_b);
  target += beta_lpdf(tau | 2, 8); 

  V[1] = 0.5;
  target += bernoulli_lpmf(h[1] | 0.5);

  for (t in 2:n) {
    V[t] = V[t-1] + alpha * (feedback[t-1] - V[t-1]);
    real p = 1 / (1 + exp(-tau_scaled * V[t]));
    target += bernoulli_lpmf(h[t] | p);
  }
}

generated quantities {

  // Prior samples
  real<lower=0, upper=1> alpha_prior = beta_rng(prior_alpha_a, prior_alpha_b);
  real<lower=0, upper=1> tau_prior   = beta_rng(2, 8); 
  real<lower=0, upper=20> tau_prior_scaled = tau_prior * 20;

  // Prior Predictive Check 
  array[n] real V_prior; // stores V at each trial while computing prior simulation
  array[n] int  h_prior_rep; // stores simulated choices while computing prior simulation

  V_prior[1] = 0.5;
  h_prior_rep[1] = bernoulli_rng(0.5);

  for (t in 2:n) {
    V_prior[t] = V_prior[t-1] + alpha_prior * (feedback[t-1] - V_prior[t-1]);
    real p_prior = 1 / (1 + exp(-tau_prior_scaled * V_prior[t]));
    h_prior_rep[t] = bernoulli_rng(p_prior);
  }

  // total right choices from prior simulation (used for histo)
  int prior_rep_sum = sum(h_prior_rep); 

  // Posterior Predictive Check 
  // same structure, but now with the posterior values of alpha and tau
  array[n] real V_post;
  array[n] int  h_post_rep;

  V_post[1] = 0.5;
  h_post_rep[1] = bernoulli_rng(0.5);

  for (t in 2:n) {
    V_post[t] = V_post[t-1] + alpha * (feedback[t-1] - V_post[t-1]);
    real p_post = 1 / (1 + exp(-tau_scaled * V_post[t]));
    h_post_rep[t] = bernoulli_rng(p_post);
  }

  int post_rep_sum = sum(h_post_rep);

  // Log-likelihood
  array[n] real log_lik;
  log_lik[1] = bernoulli_lpmf(h[1] | 0.5);
  for (t in 2:n) {
    real V_ll;
    V_ll = 0.5;
    for (s in 2:t) {
      V_ll = V_ll + alpha * (feedback[s-1] - V_ll);
    }
    real p_ll = 1 / (1 + exp(-tau_scaled * V_ll));
    log_lik[t] = bernoulli_lpmf(h[t] | p_ll);
  }
}
"
```

## creating shifting bias agent and fitting model
```{r}
set.seed(42)

# 1. Shifting opponent
trials            <- 500
shifting_opponent <- c(
  rbinom(250, 1, 0.8),
  rbinom(250, 1, 0.2)
)

# 2. Simulate agent data against shifting opponent
alpha <- 0.8
tau   <- 1

df <- simulate_rl_data(
  trials            = trials,
  alpha             = alpha,
  tau               = tau,
  opponents_choices = shifting_opponent
)

# 3. Data for Stan
data_for_stan <- list(
  n             = trials,
  h             = df$choice_numeric,
  h_opponent    = shifting_opponent,
  feedback      = shifting_opponent,
  prior_alpha_a = 2,
  prior_alpha_b = 2
)

# 4. Write Stan model to file
stan_model_dir <- "~/Documents/GitHub/CogSci_MSc2/advanced_cognitive_modeling/Assignment_2/stan"
stan_file_RL   <- here(stan_model_dir, "RL_model.stan")
if (!dir.exists(stan_model_dir)) dir.create(stan_model_dir, recursive = TRUE)
writeLines(stan_model_code, stan_file_RL)

# 5. Compile and fit
stan_results_dir <- "~/Documents/GitHub/CogSci_MSc2/advanced_cognitive_modeling/Assignment_2/simmodels"
model_file_RL    <- file.path(stan_results_dir, "RL_model1.rds")
if (!dir.exists(stan_results_dir)) dir.create(stan_results_dir, recursive = TRUE)

mod_RL <- cmdstan_model(
  stan_file_RL,
  cpp_options   = list(stan_threads = FALSE),
  stanc_options = list("O1")
)

fit_RL <- mod_RL$sample(
  data            = data_for_stan,
  seed            = 123,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1000,
  iter_sampling   = 2000,
  refresh         = 0,
  max_treedepth   = 10,
  adapt_delta     = 0.8
)

fit_RL$save_object(file = model_file_RL)
```
# model diagnostics
```{r}
# Extract the summary and diagnostics from our fitted model object
fit_summary <- fit_RL$summary()
diagnostic_summary <- fit_RL$diagnostic_summary()

# 1. Check for Divergent Transitions (Zero Tolerance)
divergences <- sum(diagnostic_summary$num_divergent)
if (divergences > 0) {
  stop(paste("CRITICAL FAILURE:", divergences, "divergences detected. The posterior geometry is pathological. Reparameterization required. Do not interpret these results."))
}

# 2. Check E-BFMI (Energy-Bayesian Fraction of Missing Information)
# Threshold is typically 0.3. Values below this indicate poor exploration.
min_ebfmi <- min(diagnostic_summary$ebfmi)
if (min_ebfmi < 0.3) {
  warning(paste("WARNING: Low E-BFMI detected (Min =", round(min_ebfmi, 3), 
                "). The sampler may struggle to explore the tails of the posterior."))
}

# 3. Check for Convergence (Rhat < 1.01)
max_rhat <- max(fit_summary$rhat, na.rm = TRUE)
if (max_rhat > 1.01) {
  stop(paste("CRITICAL FAILURE: Chains did not converge. Max Rhat =", round(max_rhat, 3)))
}

# 4. Check Effective Sample Size (ESS > 400 for both bulk and tail)
min_ess_bulk <- min(fit_summary$ess_bulk, na.rm = TRUE)
min_ess_tail <- min(fit_summary$ess_tail, na.rm = TRUE)
if (min_ess_bulk < 400 || min_ess_tail < 400) {
  warning("WARNING: Insufficient Effective Sample Size. Autocorrelation is too high. Run longer chains.")
}

# 5. Print summary for key parameters only
fit_RL$summary(variables = c("alpha", "tau", "tau_scaled", "alpha_prior", "tau_prior"))
```


# Plot for prior predictive check
```{r}
# Extract the summary statistic calculated from prior predictive simulations
# 'prior_rep_sum' was calculated within the generated quantities block
draws_gq <- as_draws_df(fit_RL$draws()) # Ensure we have the draws as a dataframe
observed_sum <- sum(data_for_stan$h) # Total number of 'right' choices in the actual observed data

ppc_prior_ggplot <- ggplot(draws_gq, aes(x = prior_rep_sum)) +
  # Draw the histogram of the model's prior predictions (light red)
  geom_histogram(fill = "#ffbaba", color = "#ff5252", bins = 30, alpha = 0.8) +
  
  # Add the vertical line for the actual observed data (dark red)
  geom_vline(xintercept = observed_sum, color = "darkred", linewidth = 1.2) +
  
  # Labels and theming
  ggtitle("Prior Predictive Check: Total 'Right' Choices") +
  xlab(paste("Total Number of 'Right' Choices (out of", trials, "trials)")) +
  ylab("Frequency") +
  theme_classic()

ppc_prior_ggplot
```

# Combined prior and posterior predictive check plot
```{r}
draws_gq <- as_draws_df(fit_RL$draws())

predictive_df <- tibble(
  Value = c(draws_gq$prior_rep_sum, draws_gq$post_rep_sum),
  Type  = rep(c("Prior Predictive", "Posterior Predictive"), each = nrow(draws_gq))
)

ggplot(predictive_df, aes(x = Value, fill = Type)) +
  geom_histogram(alpha = 0.6, color = "white", bins = 30) +
  geom_vline(xintercept = sum(data_for_stan$h),
             linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c(
    "Prior Predictive"     = "gray70",
    "Posterior Predictive" = "skyblue"
  )) +
  labs(
    title = "Predictive Check: Total 'Right' Choices",
    x     = paste("Total Number of 'Right' Choices (out of", trials, "trials)"),
    y     = "Frequency"
  ) +
  theme_classic()
```
-----------------------

# Prior sensitivity
```{r}
#defining a grid of prior values to try - only for alpha right now, and since it is a beta prior there is an alpha and a beta parameter to vary
prior_grid <- tidyr::expand_grid(
  prior_a = c(1, 2, 5, 10, 20), 
  prior_b = c(1, 2, 5, 10, 20)
)


# Standardized filename variable
sensitivity_draws_file <- file.path(here(stan_results_dir), "W5_sensitivity_draws.rds")

if (regenerate_simulations || !file.exists(sensitivity_draws_file)) {
  # Removed 'seed_val' from the function arguments
  fit_and_extract <- function(prior_a, prior_b) {
    # Dynamically inject the perturbed parameters into the data list
    current_data_list <- data_for_stan
    current_data_list$prior_alpha_a <- prior_a
    current_data_list$prior_alpha_b <- prior_b

    # Fit using the original compiled model
    fit <- mod_RL$sample(
      data = current_data_list,
      chains = 1,
      parallel_chains = 1, 
      iter_warmup = 1000,
      iter_sampling = 3000,
      refresh = 0, 
      adapt_delta = 0.8,
      show_messages = FALSE,
      show_exceptions = FALSE # Added to prevent C++ errors from polluting stdout
    )
    
    draws_df <- fit$draws("alpha", format = "df")
    
    # Return a strictly structured tibble, discarding the fragile fit object
    tibble::tibble(
      alpha_prob = draws_df$alpha,
      prior_a = prior_a,
      prior_b = prior_b
    )
  }

  # Establish the parallel backend 
  future::plan(future::multisession, workers = 2)
  
  # Execute the parallel mapping
  draws_list <- furrr::future_pmap(
    prior_grid,
    fit_and_extract,
    .options = furrr::furrr_options(seed = TRUE, stdout = FALSE) 
  )
  
  # Collapse the list of dataframes into a single tidy dataframe
  sensitivity_draws <- dplyr::bind_rows(draws_list)
  
  # Shut down the parallel backend to free up RAM
  future::plan(future::sequential)
  
  # Save the indestructible, pure data (fixed variable name)
  saveRDS(sensitivity_draws, sensitivity_draws_file)
  
} else {
  # Fixed variable names to match what was saved
  sensitivity_draws <- readRDS(sensitivity_draws_file)
  cat("Loaded existing perturbation grid from:", sensitivity_draws_file, "\n")
}

```

```{r}
# saving the true alpha for the visualization
true_alpha= 0.8
```


```{r}
#visualizing the prior sensitivity code from riccardo
p_sensitivity_2d <- ggplot(sensitivity_draws, aes(x = alpha_prob, group = factor(prior_b), colour = prior_b)) +
  geom_line(stat = "density", alpha = 0.8, linewidth = 0.8) + 
  geom_vline(xintercept = true_alpha, color = "black", linetype = "dashed", linewidth = 1) + 
  facet_wrap(~ prior_a, labeller = labeller(prior_a = function(x) paste0("Alpha prior a parameter: ", x))) +
  scale_color_viridis_c(option = "mako", name = expression(Prior ~ sigma)) + 
  coord_cartesian(xlim = c(0.4, 1.0)) + 
  ggtitle("Epistemic Robustness: 2D Prior Perturbation Grid") +
  xlab(expression(theta ~ " (Probability scale)")) +
  ylab("Marginal Density") +
  theme_classic() +
  theme(
    legend.position = "bottom",
    strip.background = element_rect(fill = "gray95", color = "NA"),
    strip.text = element_text(face = "bold", size = 11)
  )

p_sensitivity_2d
```
For some reason the impact of the prior is huge now - not good for the model but perhaps more indicative that the test is working

# Prior sensitvity analysis - trajectory

This plot shows how the posterior mean of alpha and its credible intervals change when different priors are used for alpha.
```{r}
sensitivity_summary <- sensitivity_draws |> 
  dplyr::group_by(prior_a, prior_b) |> 
  dplyr::summarize(
    post_mean = mean(alpha_prob),
    lower_95 = quantile(alpha_prob, 0.025),
    upper_95 = quantile(alpha_prob, 0.975),
    lower_68 = quantile(alpha_prob, 0.16),
    upper_68 = quantile(alpha_prob, 0.84),
    .groups = "drop"
  ) |> 
  dplyr::mutate(
    prior_a_label = factor(
      prior_a, 
      levels = c(1, 2, 5, 10, 20), 
      labels = c(
        "Prior a = 1",
        "Prior a = 2",
        "Prior a = 5",
        "Prior a = 10",
        "Prior a = 20"
      )
    )
  )

p_sensitivity_trajectory <- ggplot(
  sensitivity_summary,
  aes(x = prior_b, y = post_mean, group = 1)
) +
  geom_hline(yintercept = true_alpha, color = "black", linetype = "dashed", linewidth = 1) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), alpha = 0.15, fill = "steelblue") +
  geom_ribbon(aes(ymin = lower_68, ymax = upper_68), alpha = 0.35, fill = "steelblue") +
  geom_line(linewidth = 1, color = "steelblue4") +
  geom_point(size = 2, color = "steelblue4") +
  facet_wrap(~ prior_a_label) +
  coord_cartesian(ylim = c(0, 1)) +
  ggtitle("Prior Sensitivity Summary for Alpha") +
  xlab("Beta prior b parameter") +
  ylab("Posterior estimate of alpha (mean, 68%, 95% CrI)") +
  theme_classic() +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "gray95", color = NA),
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.major.y = element_line(color = "gray90", linetype = "dotted")
  )

print(p_sensitivity_trajectory)
```
- If the model were not very sensitive to the prior, then the blue line would stay almost flat and close to the black dashed line.
- increasing the Beta prior parameter b pulls the posterior estimate downward across all values of a
- Uncertainty becomes smaller for larger values of a
- Dots = posterior mean


#### parameter recovery

```{r}
#code for testing how the simulate_rl_data function works before implementing it in the big loop below
#n_trials = c(100)
#true_alpha = c(0.5)
#true_tau = c(1) # fixed temperature until the model works
#sim_opp <- rbinom(n = n_trials, size = 1, prob = 0.8)
#sim_h <- simulate_rl_data(trials = n_trials, alpha = true_alpha, tau = true_tau, opponents_choices = sim_opp) # outputs the last element in the function - which is the choices as numeric

```


```{r}


recovery_grid <- tidyr::expand_grid(
  n_trials = c(100, 200, 300, 400), #varying ntrials
  true_alpha = c(0.5, 0.7, 0.9), # same alpha variation as the theta variation in course notes
  true_tau = c(1) # fixed temperature until the model works
)

recovery_file <- file.path(stan_results_dir, "W5_recovery_draws.rds")


if (regenerate_simulations || !file.exists(recovery_file)) {
  
  # 2. Define the isolated recovery pipeline
  run_recovery <- function(n_trials, true_alpha, true_tau) {
    # Generate synthetic empirical data
    sim_opp <- rbinom(n = n_trials, size = 1, prob = 0.8) # Simulating the opponent THIS MIGHT NEED TO CHANGE TO MEET RICCARDOS SUGGESTION
    sim_h <- simulate_rl_data(trials = n_trials, alpha = true_alpha, tau = true_tau, opponents_choices = sim_opp) # using our function
    
    
    # Structure the data block for Stan
    rec_data_list <- list(
      n = n_trials,
      h_opponent = sim_opp,
      h = sim_h$choice_numeric,
      feedback = sim_h$reward_right, 
      prior_alpha_a = 2,
      prior_alpha_b = 2
    )
    
    # Fit the generative model
    fit <- mod_RL$sample(
      data = rec_data_list,
      chains = 1,
      iter_warmup = 1000,
      iter_sampling = 2000,
      refresh = 0,
      show_messages = FALSE
    )
    
    # Extract structural draws immediately to avoid temporary file fragility
    draws_df <- fit$draws("alpha", format = "df")
    
    tibble::tibble(
      trials = n_trials,
      true_alpha = true_alpha,
      true_tau = true_tau,
      alpha = draws_df$alpha
    )
  }

  # 3. Establish the parallel backend and execute mapping
  future::plan(future::multisession, workers = 2)
  
  recovery_draws <- furrr::future_pmap(
    recovery_grid,
    run_recovery,
    .options = furrr::furrr_options(seed = TRUE, stdout = FALSE)
  ) |> 
    dplyr::bind_rows()
    
  future::plan(future::sequential)
  
  saveRDS(recovery_draws, recovery_file)
  
} else {
  recovery_draws <- readRDS(recovery_file)
  cat("Loaded existing 2D recovery draws from:", recovery_file, "\n")
}
```

```{r}
# plotting the paramenter revcovery MIGHT NEED ADJSUTMENT OF AXES
# 4. Compress draws into rigorous posterior summaries
recovery_summary <- recovery_draws |> 
  dplyr::group_by(trials, true_alpha) |> 
  dplyr::summarize(
    post_mean = mean(alpha),
    lower_95 = quantile(alpha, 0.025),
    upper_95 = quantile(alpha, 0.975),
    .groups = "drop"
  ) |> 
  dplyr::mutate(
    facet_label = paste0("Ground Truth: \u03b8 = ", true_alpha)
  )

# 5. Visualize the geometric shrinkage across the parameter space

p_recovery_2d <- ggplot(recovery_summary, aes(x = factor(trials), y = post_mean)) +
  # Map the true value to the hline dynamically per facet
  geom_hline(aes(yintercept = true_alpha), color = "black", linetype = "dashed", linewidth = 1) +
  geom_pointrange(aes(ymin = lower_95, ymax = upper_95, color = factor(true_alpha)), linewidth = 1, size = 0.8) +
  
  facet_wrap(~ facet_label) +
  # Adjust palette range so colors remain distinct
  scale_color_viridis_d(option = "mako", begin = 0.2, end = 0.8) +
  
  ggtitle("Data Adequacy & Identifiability: Parameter Recovery") +
  xlab("Experimental Design (Number of Trials, N)") +
  ylab(expression("Posterior Estimate of " * alpha * " (Mean & 95% CrI)")) +
  
  coord_cartesian(ylim = c(0.35, 1.0)) + # Bound visually to the relevant domain
  theme_classic() +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "gray95", color = "NA"),
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.major.y = element_line(color = "gray90", linetype = "dotted")
  )

print(p_recovery_2d)
```



```{r}
#plotting an alternative visualization of the recovery

p_recovery_2d_alt <- ggplot(recovery_summary, aes(x = true_alpha, y = post_mean)) +
  # The Identity Line: y = x (Estimate perfectly matches Ground Truth)
  geom_abline(intercept = 0, slope = 1, color = "black", linetype = "dashed", linewidth = 1) +
  
  # The Posterior Estimates and 95% CrI
  geom_pointrange(aes(ymin = lower_95, ymax = upper_95, color = factor(true_alpha)), linewidth = 1, size = 0.8) +
  
  # Facet strictly by the Experimental Design (N trials)
  facet_wrap(
    ~ trials, 
    labeller = as_labeller(function(x) paste0(x, " Trials"))
  ) +
  
  scale_color_viridis_d(option = "mako", begin = 0.2, end = 0.8) +
  
  # Ensure the coordinate space is perfectly square to preserve the visual geometry of y=x
  coord_fixed(xlim = c(0.4, 1.0), ylim = c(0.4, 1.0)) + 
  
  xlab(expression("Generative Ground Truth (" * alpha * ")")) +
  ylab(expression("Posterior Estimate (" * hat(alpha) * " Mean & 95% CrI)")) +
  
  theme_classic() +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "gray95", color = "NA"),
    strip.text = element_text(face = "bold", size = 11),
    # Add a faint grid to help the eye assess the distance from the diagonal
    panel.grid.major = element_line(color = "gray90", linetype = "dotted")
  )

print(p_recovery_2d_alt)
```
The number of trials doesn't really seem to teach us much

